---
title: "Statistical and Machine Learning"
subtitle: "Lab2: Regularization <br> Ridge, LASSO, and Elastic-net Regression"
author: "Tsai, Dai-Rong"
format:
  revealjs:
    theme: default
    echo: true
    smaller: true
    scrollable: true
    slide-number: true
    auto-stretch: false
    history: false
    pdf-max-pages-per-slide: 5
    embed-resources: true
tbl-cap-location: bottom
---

## Dataset

```{r echo = FALSE}
options(digits = 5, width = 100)
```

```{css echo = FALSE}
.reveal table, ul ul li, p code{
  font-size: smaller;
}
```

> ***Prostate Cancer Data***

```{r}
# Set random seed
set.seed(1234)

# Packages
library(MASS) # for stepAIC
library(glmnet) # for glmnet, glmnet.cv
library(sparsegl) # for sparsegl, cv.sparsegl

# Data
prostate <- read.table("http://www.stat.cmu.edu/~ryantibs/statcomp/data/pros.dat")
```

- Response
    - `lpsa`: log(prostate-specific antigen)
- Predictors
    - `lcavol`: log(cancer volume)
    - `lweight`: log(prostate weight)
    - `age`: age
    - `lbph`: log(amount of benign prostatic hyperplasia)
    - `svi`: seminal vesicle invasion
    - `lcp`: log(capsular penetration)
    - `gleason`: Gleason scores
    - `pgg45`: percentage Gleason scores 4 or 5

---

::: {.panel-tabset}

### Preview

```{r}
dim(prostate)
head(prostate)
```

### Data Structure

```{r}
str(prostate)
```

:::

## Create Training/Testing Partitions

- Split data into 80% training set and 20% test set

```{r}
nr <- nrow(prostate)
train.id <- sample(nr, nr * 0.8)

training <- prostate[train.id, ]
testing <- prostate[-train.id, ]
```

- Check dimension

```{r}
dim(training)
dim(testing)
```

## Model Formulae in R

| Formula | Model |
|--------|--------|
| `Y ~ 1` | $Y = \beta_0 + \epsilon$ |
| `Y ~ X1 + X2` | $Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \epsilon$ |
| `Y ~ X1 + X2 - 1` | $Y = \beta_1X_1 + \beta_2X_2 + \epsilon$ |
| `Y ~ .` | $Y = \beta_0 + \beta_1X_1 + \ldots + \beta_pX_p + \epsilon$ |
| `Y ~ . - Xp` | $Y = \beta_0 + \beta_1X_1 + \ldots + \beta_{p-1}X_{p-1} + \epsilon$ |
| `Y ~ X1:X2` | $Y = \beta_0 + \beta_{12}X_1X_2 + \epsilon$ |
| `Y ~ X1 * X2` | $Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_{12}X_1X_2 + \epsilon$ |
| `Y ~ (X1 + X2 + X3)^2` | $\begin{aligned}
                            Y = &\beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \\
                                &\beta_{12}X_1X_2 + \beta_{13}X_1X_3 + \beta_{23}X_2X_3 + \epsilon \\
                                &(\color{red}{\beta_{123}X_1X_2X_3} \text{ is NOT contained})
                            \end{aligned}$ |
| `Y ~ X1 + I(X1^2)` <br> `Y ~ poly(X1, 2, raw = TRUE)` | $Y = \beta_0 + \beta_1X_1 + \beta_2X_1^2 + \epsilon$ |

: Model Formulae in R {#tbl-formula}

## Subset Selection

::: {.callout-note}

### `stats::step()`

- This is a minimal implementation. Use `stepAIC` in package `MASS` for a wider range of object classes.
- B. D. Ripley: `step` is a slightly simplified version of `stepAIC` in package `MASS`.

:::

```{r}
lm.full <- lm(lpsa ~ ., data = training)
```

:::: {.columns}

::: {.column width="50%"}

- Forward Selection

```{r}
lm.for <- stepAIC(update(lm.full, . ~ 1),
                  scope = list(upper = lm.full),
                  direction = "forward",
                  trace = 0)
lm.for$anova
```

:::

::: {.column width="50%"}

- Backward Selection

```{r}
lm.back <- stepAIC(lm.full,
                   scope = list(lower = ~ 1),
                   direction = "backward",
                   trace = 0)
lm.back$anova
```
:::

::::

- Stepwise Selection

```{r}
lm.step <- stepAIC(lm.full, scope = list(lower = ~ 1), direction = "both", trace = 0)
lm.step$anova
```

## Regularization

- Construct Design Matrix

```{r}
x <- model.matrix(lpsa ~ ., data = training)[, -1]
y <- training$lpsa
```

::: {.callout-note appearance="minimal"}

- `"lpsa ~ ."`: Create a design matrix for all variables except for `lpsa`.
- `"[, -1]"`: Exclude the column of intercepts. (The intercept term is fitted by default in `glmnet`)

:::

### Ridge / Lasso Regression

```{r}
ridge <- glmnet(x = x, y = y, alpha = 0) # Ridge
lasso <- glmnet(x = x, y = y, alpha = 1) # Lasso
```

::: {.callout-tip}

### Arguments

- `family`:
    - One of the built-in families: `"gaussian"`(default), `"binomial"`, `"poisson"`, `"multinomial"`, `"cox"`, `"mgaussian"`.
    - A `glm()` family object. (See `?family`)

:::

```{r}
#| fig-width: 10
#| fig-height: 7
#| code-fold: true
#| code-summary: "codes for plot"

par(mfrow = c(2, 2))
plot(ridge, xvar = "lambda", label = TRUE)
plot(ridge, xvar = "norm", label = TRUE)
plot(lasso, xvar = "lambda", label = TRUE)
plot(lasso, xvar = "norm", label = TRUE)
title("Ridge Regression", line = -2, outer = TRUE)
title("Lasso Regression", line = -22, outer = TRUE)
```

```{r echo = FALSE}
op <- options(digits = 2)
```

:::: {.columns}

::: {.column width="50%"}

- Coefficients of Ridge

```{r}
coef(ridge, s = exp(seq(-2, 6, 2)))
```

:::

::: {.column width="50%"}

- Coefficients of Lasso

```{r}
coef(lasso, s = exp(seq(-5, -1, 1)))
```

:::

::::

```{r echo = FALSE}
options(op)
```

---

### Cross-Validation

:::: {.columns}

::: {.column width="50%"}

- CV for Ridge

```{r}
ridge.cv <- cv.glmnet(x = x, y = y, alpha = 0,
                      type.measure = "mse",
                      nfolds = 10)
ridge.cv
plot(ridge.cv)
coef(ridge.cv, s = "lambda.min")
```

:::

::: {.column width="50%"}

- CV of Lasso

```{r}
lasso.cv <- cv.glmnet(x = x, y = y, alpha = 1,
                      type.measure = "mse",
                      nfolds = 10)
lasso.cv
plot(lasso.cv)
coef(lasso.cv, s = "lambda.min")
```

:::

::::

::: {.callout-tip}

### Arguments

- `type.measure`: loss to use for cross-validation.
    - `"default"` : MSE for gaussian models, deviance for logistic and poisson regressions, and partial-likelihood for the Cox model.
    - `"mse"`/`"mae"`: Mean squared/absolute error for all models except the "Cox".
    - `"deviance"`: Deviance for logistic and poisson regressions.
    - `"class"`: Misclassification error for binomial and multinomial logistic regressions.
    - `"auc"`: Area under the ROC curve for ***two-class*** logistic regression.
    - `"C"`: Harrel's concordance measure for ***Cox*** models.
- `s`: Value(s) of the penalty parameter $\lambda$.
    - `"lambda.1se"` (default): Largest value of $\lambda$ such that error is within 1 standard error of the minimum.
    - `"lambda.min"`: Value of $\lambda$ that gives the minimum mean cross-validated error.
    - numeric vector: Value(s) of $\lambda$ to be used

:::

---

### Elastic-Net

- CV of Elastic-Net

```{r}
foldid <- sample(cut(1:nrow(training), 10, labels = FALSE))
alpha.grid <- seq(0.1, 0.9, 0.2)

elnet.cv <- do.call(rbind,
  lapply(alpha.grid, \(a) {
    cv <- cv.glmnet(x = x, y = y, foldid = foldid, alpha = a)
    data.frame(alpha = a, cv[c("lambda", "cvm")])
  })
)
```

:::: {.columns}

::: {.column width="50%"}

```{r}
head(elnet.cv)
```

:::

::: {.column width="50%"}

```{r}
tail(elnet.cv)
```

:::

::::

::: {.callout-tip}

### Arguments

- `foldid`: an optional vector of values between 1 and `nfolds` identifying what fold each observation is in. If supplied, `nfolds` can be missing.

Users can explicitly control the fold that each observation is assigned to via the `foldid` argument.
This is useful in using cross-validation to select a value for $\alpha$.

:::

```{r}
#| fig-width: 8
#| fig-height: 4
#| code-fold: true
#| code-summary: "codes for plot"

library(ggplot2)

ggplot(elnet.cv, aes(x = log(lambda), y = cvm, colour = factor(alpha))) +
  geom_line() + geom_point(size = 0.5) +
  scale_color_viridis_d(name = quote(alpha)) +
  labs(x = quote(log(lambda)), y = "Cross-Validation MSE") +
  theme_bw()
```

- Optimal CV Parameters

```{r}
(elnet.optim <- elnet.cv[which.min(elnet.cv$cvm), ])
elnet <- glmnet(x = x, y = y, alpha = elnet.optim$alpha, lambda = elnet.optim$lambda)
coef(elnet)
```

---

### (Sparse-) Group Lasso

```{r}
(grp <- setNames(rep(1:4, each = 2), colnames(x)))

grplasso <- sparsegl(x = x, y = y, group = grp, asparse = 0)
grplasso.cv <- cv.sparsegl(x = x, y = y, group = grp, asparse = 0, pred.loss = "mse")
grplasso.cv
```

::: {.callout-tip}

### Arguments

- `group`: A vector of consecutive integers describing the grouping of the coefficients.
- `asparse`: The relative weight to put on the $\ell_1$-norm in sparse group lasso. 
    - `asparse = 0` gives the group lasso fit; `asparse = 1` gives the lasso fit.

:::

```{r}
coef(grplasso.cv, s = "lambda.1se")
```

```{r}
#| fig-width: 12
#| fig-height: 4
#| code-fold: true
#| code-summary: "codes for plot"

library(patchwork)
p1 <- plot(grplasso, y_axis = "group")
p2 <- plot(grplasso.cv)
p1 + p2
```

## Prediction

```{r}
pred.for <- predict(lm.for, testing)
pred.back <- predict(lm.back, testing)
pred.step <- predict(lm.step, testing)

z <- model.matrix(lpsa ~ ., data = testing)[, -1]
pred.ridge <- predict(ridge.cv, newx = z, s = "lambda.min")
pred.lasso <- predict(lasso.cv, newx = z, s = "lambda.min")
pred.elnet <- predict(elnet, newx = z)

mse <- sapply(mget(ls(pattern = "^pred")), \(x) mean((x - testing$lpsa)^2))
```

```{r echo = FALSE}
names(mse) <- sub("pred.", "", names(mse))
```

```{r}
sort(mse)
```

```{r echo = FALSE}
#| fig-width: 7
#| fig-height: 5

barplot(sort(mse), ylim = range(mse) + c(-0.01, 0.01), xpd = FALSE, col = 4,
        ylab = "MSE", main = "Prediction Error")
```

## References

- [`{glmnet}` Website](https://glmnet.stanford.edu/articles/glmnet.html)
- [`{sparsegl}` Website](https://dajmcdon.github.io/sparsegl/articles/sparsegl.html)
